---
title: "Figure 4.5: Te Reo Māori toponyms in Aotearoa"
categories: 
  - figures
  - code
  - R
execute: 
  cache: true
  freeze: auto
knitr:
  opts_chunk: 
    warning: false
    message: false
---

I made an alternative version of this map for some slides. It shows off the little-known multiplication by an affine transform matrix superpower that `sf` somewhat surprisingly admits (it felt like a cheat-code when I came across it [here](https://r-spatial.github.io/sf/articles/sf3.html#affine-transformations)).

```{r message=FALSE, warning=FALSE, results=FALSE}
library(sf)
library(tmap)
library(dplyr)

ang <- 40 * pi / 180
rotation <- matrix(c(cos(ang), -sin(ang), 
                     sin(ang),  cos(ang)), 2, 2, byrow = TRUE)

nz <- st_read("nz-2193.gpkg") %>%
  st_cast("POLYGON") %>%
  mutate(geom = geom * rotation)

toponyms <- st_read("placenames.gpkg") %>%
  mutate(maori = reo > 0) %>%
  mutate(geom = geom * rotation) %>%
st_filter(nz)

tm_shape(nz) + 
  tm_fill(col = "#eee8d5") + 
  tm_shape(toponyms %>% arrange(id)) +
  tm_dots(col = "maori", palette = c("slategrey", "red"), size = 0.005, alpha = 0.2,
          legend.show = FALSE) +
  tm_layout(frame = FALSE, inner.margins = 0.05, bg.color = "#002b36")
```

As noted in the book, the materials for this figure draw on the [amazing work of Chris McDowall and Tim Denee](https://github.com/fogonwater/we-are-here/) in their [_We Are Here_ atlas](https://www.masseypress.ac.nz/books/we-are-here/), the [Toitū Te Whenua &ndash; Land Information New Zealand](https://www.linz.govt.nz/) [gazetteer of placenames](https://data.linz.govt.nz/layer/51681-nz-place-names-nzgb/), and on [Te Hiku Media](https://tehiku.nz/)'s [Ngā-kupu tools](https://github.com/TeHikuMedia/nga-kupu).  

```{r}
#| code-fold: true
# License (MIT)
#
# Copyright (c) 2023 David O'Sullivan
#
# Permission is hereby granted, free of charge, to any person
# obtaining a copy of this software and associated documentation
# files (the "Software"), to deal in the Software without restriction,
# including without limitation the rights to use, copy, modify, merge,
# publish, distribute, sublicense, and/or sell copies of the Software,
# and to  permit persons to whom the Software is furnished to do so,
# subject to the following conditions:
#
# The above copyright notice and this permission notice shall be included
# in all copies or substantial portions of the Software.
#
# THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS
# OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
# FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL
# THE AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
# LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING
# FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
# DEALINGS IN THE SOFTWARE.
```
