---
title: "Figure 7.6 Space-time paths for 100 Beijing taxis"
categories: 
  - figures
  - code
  - R
execute: 
  cache: true
  freeze: auto
knitr:
  opts_chunk: 
    warning: false
    message: false
---

Taking advantage of the web format, here I show an interactive version of this figure.

```{r}
library(dplyr)
library(timeDate)
library(gg3D)
library(plotly)
```

We filter the data so they are all within 0.25 decimal degrees of the mean location of all points (I know, I know... decimal degrees are terrible, but the point here is the time dimension, not the geographical coordinates).

```{r}
taxis <- read.table('taxis100.txt', sep=',') %>%
  rename(id = V1, t = V2, lon = V3, lat = V4)

lon.mean <- mean(taxis$lon)
lat.mean <- mean(taxis$lat)

taxis.bj <- taxis %>%
  mutate(time = as.double(timeDate(t))) %>%
  filter(abs(lon - lon.mean) < .25,
         abs(lat - lat.mean) < .25)
```

`plotly` makes a nice interactive plot, without too much fuss.

```{r fig.width=10, fig.height=8}
plot_ly(group_by(taxis.bj, id), 
        x = ~lat, y = ~lon, z = ~t, color = ~id, 
        type = 'scatter3d', mode = 'lines', asp = 1, lwd = 0.5)
```

And more like the plot in the book, here is a static 2.5D version, using `gg3d::stat_3D`.

```{r fig.width=10, fig.height=8}
ggplot(taxis.bj, aes(x = lon, y = lat, z = time, group = id, color = id)) +
  stat_3D(theta = 30, phi = 5, geom = "path", 
          alpha = 0.25, linewidth = 0.35) +
  scale_color_viridis_c() +
  axes_3D(theta = 30, phi = 5) +
  coord_equal() +
  theme_void() +
  theme(legend.position = "none")
```

```{r}
#| code-fold: true
# License (MIT)
#
# Copyright (c) 2023 David O'Sullivan
#
# Permission is hereby granted, free of charge, to any person
# obtaining a copy of this software and associated documentation
# files (the "Software"), to deal in the Software without restriction,
# including without limitation the rights to use, copy, modify, merge,
# publish, distribute, sublicense, and/or sell copies of the Software,
# and to  permit persons to whom the Software is furnished to do so,
# subject to the following conditions:
#
# The above copyright notice and this permission notice shall be included
# in all copies or substantial portions of the Software.
#
# THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS
# OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
# FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL
# THE AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
# LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING
# FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
# DEALINGS IN THE SOFTWARE.
```
