---
title: "Figure 7.7 Coordination in space-time paths"
categories: 
  - figures
  - code
  - R
execute: 
  cache: true
  freeze: auto
knitr:
  opts_chunk: 
    warning: false
    message: false
---

This page shows a different set of 'flockers' converging in space-time&mdash;in fact the same ones that are used on the ['make your own cover' page](../../other-stuff/about-the-cover.qmd). I've provided two plots here, both in colour, and one an interactive 3D plot so you can have a proper look.

```{r}
library(dplyr)
library(gg3D)
library(plotly)

flockers <- read.table('traces.csv', sep=' ', header = TRUE)
```

It's worth noting that `plotly` rescales all the coordinates here (because I ask it to with `asp = 1`) but that's probably best for giving a general impression of the patterns.

```{r fig.width=10, fig.height=8}
plot_ly(group_by(flockers, who), 
        x = ~x, y = ~y, z = ~t, color = ~who, colors = "magma",
        type = 'scatter3d', mode = 'lines', asp = 1, lwd = 0.5)
```

Not nearly as much fun is the static plot (like the "library in a theme park", as [my favourite cricketer once said](https://www.stuff.co.nz/sport/cricket/64593956/kane-williamson-says-batting-with-brendon-mccullum-like-being-library-in-a-theme-park)).

```{r fig.width=10, fig.height=8}
ggplot(flockers, aes(x = x, y = y, z = t, group = who, color = who)) +
  stat_3D(theta = 135, phi = 35, geom = "path",
          linewidth = 0.35, alpha = 0.75) +
  scale_color_viridis_c(option = "A") +
  coord_equal() +
  theme_void() +
  theme(legend.position = "none")
```

```{r}
#| code-fold: true
# License (MIT)
#
# Copyright (c) 2023 David O'Sullivan
#
# Permission is hereby granted, free of charge, to any person
# obtaining a copy of this software and associated documentation
# files (the "Software"), to deal in the Software without restriction,
# including without limitation the rights to use, copy, modify, merge,
# publish, distribute, sublicense, and/or sell copies of the Software,
# and to  permit persons to whom the Software is furnished to do so,
# subject to the following conditions:
#
# The above copyright notice and this permission notice shall be included
# in all copies or substantial portions of the Software.
#
# THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS
# OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
# FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL
# THE AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
# LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING
# FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
# DEALINGS IN THE SOFTWARE.
```
