---
title: "Figure 2.4: Two point patterns in absolute space"
execute: 
  cache: true
  freeze: auto
---

Although it is labelled 'point patterns in absolute space', the real (ahem) _point_ of this figure is that spatial analysis often works, in effect, with a relative concept of space, since it is based on notions as here, such as the distances between points.

If you want point patterns, you need [the fantastic `spatstat`](https://github.com/spatstat) by Adrian Baddeley, Ege Rubak, and Rolf Turner.

```{r message = FALSE}
library(spatstat)
```

Make a couple of point patterns, one evenly-space using _sequential spatial inhibition_, and one clustered using a _Thomas process_. For the latter, see

> Thomas M. 1949. A generalisation of Poisson's binomial limit for use in ecology. _Biometrika_ **36**, 18–25.

```{r}
pp1 <- rSSI(0.07, n = 105)
pp2 <- rThomas(10, 0.03, 10)
```

Calculate nearest neighbour distances.

```{r}
nn1 <- nndist(pp1)
nn2 <- nndist(pp2)
breaks <- seq(0, max(nn2), length.out = 10)
```

And plot them.

```{r}
par(mfrow = c(2, 2), mai = rep(0.3, 4))

plot(pp1, main = "")
hist(nn1, main = "", xlab = "Distance", ylab = "Frequency", breaks = breaks)
plot(pp2, main = "")
hist(nn2, main = "", xlab = "Distance", ylab = "Frequency", breaks = breaks)
```



```{r}
#| code-fold: true
# License (MIT)
#
# Copyright (c) 2023 David O'Sullivan
#
# Permission is hereby granted, free of charge, to any person
# obtaining a copy of this software and associated documentation
# files (the "Software"), to deal in the Software without restriction,
# including without limitation the rights to use, copy, modify, merge,
# publish, distribute, sublicense, and/or sell copies of the Software,
# and to  permit persons to whom the Software is furnished to do so,
# subject to the following conditions:
#
# The above copyright notice and this permission notice shall be included
# in all copies or substantial portions of the Software.
#
# THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS
# OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
# FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL
# THE AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
# LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING
# FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
# DEALINGS IN THE SOFTWARE.
```