---
title: "Figure 2.9 A range of spatial weights applied to polygon data"
lightbox: true
categories: 
  - figures
  - code
  - R
execute: 
  cache: true
  freeze: auto
knitr:
  opts_chunk: 
    warning: false
    message: false
---

This figure is intended to demonstrate the diversity of possible conceptualisations of 'neighbour' that are often deployed in spatial analysis contexts, and represent possible different approaches to relative space. 

The [`spdep` package](https://cran.r-project.org/web/packages/spdep/index.html) I am using here is not the easiest to use. The best guidance on `spdep` I've found is in 

> Bivand R, EJ Pebesma, and V Gómez-Rubio. 2013. _Applied Spatial Data Analysis with R_. 2nd edn. Springer.

The newer [`sfdep` package](https://cran.r-project.org/package=sfdep) is still finding its feet at time of writing (Oct 2023), but will likely be a better choice before long.

```{r message=FALSE, warning=FALSE}
library(sf)
library(dplyr)
library(spdep)
library(sp)
```

## Assembling the data
First read a polygons dataset. You'll need to supply your own for this.

```{r message=FALSE, warning=FALSE, results=FALSE}
chch <- st_read("chch-sa2.gpkg") # you need a spatial dataset
```

`spdep` requires the data to be in the `sp` package format, so we convert to that.

```{r}
polys <- chch %>%
  select(geom) %>%
  as("Spatial")
```

Now make some points inside the polygons, and also some centroids.

```{r warning=FALSE}
# guaranteed inside the polygons
pts <- chch %>%
  st_point_on_surface() %>% 
  st_geometry()

# not guaranteed, but better(?) for distance measurements
pts_c <- chch %>%
  st_centroid() %>% 
  st_geometry()
```

## Maps of adjacencies based on different rules
These are presented in a single figure in the book in four rows of three. Here I show them as sets of three for greater clarity. 

### Contiguity based

```{r}
#| fig.dim=c(8, 3)
layout(matrix(1:3, ncol = 3, byrow = TRUE))
par(mai = c(0, 0, 0.15, 0))

nb <- polys %>% poly2nb(queen = TRUE)
plot(polys, col = "lightgrey", lwd = 0.5, border = 'white', 
     main = "Queen's rule adjacency")
plot(nb, pts, col = 'red', lwd = 0.5, add = TRUE)

nb <- polys %>% poly2nb(queen = FALSE)
plot(polys, col = "lightgrey", lwd = 0.5, border = 'white', 
     main = "Rook's rule adjacency")
plot(nb, pts, col = 'red', lwd = 0.5, add = TRUE)

nb <- nb %>% nblag(2) %>% nblag_cumul()
plot(polys, col = "lightgrey", lwd = 0.5, border = 'white', 
     main = "Cumulative lag-2 adjacency")
plot(nb, pts, col = 'red', lwd = 0.5, add = TRUE)
```

### _k_-nearest neighbours
Note that we use the centroids (`pts_c`) to calculate the distances, but the points inside the polygons (`pts`) from `st_point_on_surface()` for the plotting.

```{r}
#| fig.dim=c(8, 3)
layout(matrix(1:3, ncol = 3, byrow = TRUE))
par(mai = c(0, 0, 0.15, 0))

nb <- pts_c %>% knearneigh(k = 3) %>% knn2nb()
plot(polys, col = "lightgrey", lwd = 0.5, border = 'white', 
     main = "k = 3")
plot(nb, pts, col = 'red', lwd = 0.5, add = TRUE)

nb <- pts_c %>% knearneigh(k = 6) %>% knn2nb()
plot(polys, col = "lightgrey", lwd = 0.5, border = 'white', 
     main = "k = 6")
plot(nb, pts, col = 'red', lwd = 0.5, add = TRUE)

nb <- pts_c %>% knearneigh(k = 12) %>% knn2nb()
plot(polys, col = "lightgrey", lwd = 0.5, border = 'white', 
     main = "k = 12")
plot(nb, pts, col = 'red', lwd = 0.5, add = TRUE)
```

### Distance criteria
Next, distance criteria, again calculated from centroids, but visualised using the `st_point_on_surface()`.

```{r}
#| fig.dim=c(8, 3)
layout(matrix(1:3, ncol = 3, byrow = TRUE))
par(mai = c(0, 0, 0.15, 0))

nb <- pts_c %>% dnearneigh(d1 = 0, d2 = 1000)
plot(polys, col = "lightgrey", lwd = 0.5, border = 'white', 
     main = "Distance < 1000")
plot(nb, pts, col = 'red', lwd = 0.5, add = TRUE)

nb <- pts_c %>% dnearneigh(d1 = 0, d2 = 1500)
plot(polys, col = "lightgrey", lwd = 0.5, border = 'white', 
     main = "Distance < 1500")
plot(nb, pts, col = 'red', lwd = 0.5, add = TRUE)

nb <- pts_c %>% dnearneigh(d1 = 1500, d2 = 2000)
plot(polys, col = "lightgrey", lwd = 0.5, border = 'white', 
     main = "1500 < Distance < 2000")
plot(nb, pts, col = 'red', lwd = 0.5, add = TRUE)
```

### Graph-based approaches
Finally, some network-based possibilities, [Delaunay triangulation](https://en.wikipedia.org/wiki/Delaunay_triangulation), [Gabriel graph](https://en.wikipedia.org/wiki/Gabriel_graph) and the [relative neighbour graph](https://en.wikipedia.org/wiki/Relative_neighborhood_graph).

```{r}
#| fig.dim=c(8, 3)
layout(matrix(1:3, ncol = 3, byrow = TRUE))
par(mai = c(0, 0, 0.15, 0))

g <- tri2nb(pts_c)
plot(polys, col = "lightgrey", lwd = 0.5, border = 'white', 
     main = "Delaunay triangulation")
plot(g, pts, col = 'red', lwd = 0.5, add = TRUE)

g <- gabrielneigh(pts_c)
nb <- graph2nb(g)
plot(polys, col = "lightgrey", lwd = 0.5, border = 'white', 
     main = "Gabriel graph")
plot(nb, pts, col = 'red', lwd = 0.5, add = TRUE)

g <- relativeneigh(pts_c)
nb <- graph2nb(g)
plot(polys, col = "lightgrey", lwd = 0.5, border = 'white', 
     main = "Relative neighbour graph")
plot(nb, pts, col = 'red', lwd = 0.5, add = TRUE)
```



```{r}
#| code-fold: true
# License (MIT)
#
# Copyright (c) 2023 David O'Sullivan
#
# Permission is hereby granted, free of charge, to any person
# obtaining a copy of this software and associated documentation
# files (the "Software"), to deal in the Software without restriction,
# including without limitation the rights to use, copy, modify, merge,
# publish, distribute, sublicense, and/or sell copies of the Software,
# and to  permit persons to whom the Software is furnished to do so,
# subject to the following conditions:
#
# The above copyright notice and this permission notice shall be included
# in all copies or substantial portions of the Software.
#
# THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS
# OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
# FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL
# THE AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
# LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING
# FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
# DEALINGS IN THE SOFTWARE.
```