---
title: "Figures 2.6 Slivers and gaps in a polygon layer"
categories: 
  - code
  - R
  - extra
execute: 
  cache: true
  freeze: auto
knitr:
  opts_chunk: 
    warning: false
    message: false
---

Here is how slivers and gaps can happen...

Read in some polygons.

```{r results=FALSE}
library(sf)
library(tmap)
library(dplyr)

polygons <- st_read("some-polygons.gpkg")
```

Simplify them without due care and attention, and _voila_! 

```{r fig.width=8, fig.height=5}
m1 <- tm_shape(polygons) +
  tm_fill(col = "name", alpha = 0.5) +
  tm_layout(
    title = "Source polygons", frame = FALSE,
    legend.show = FALSE)

m2 <- tm_shape(polygons %>% 
                 st_simplify(dTolerance = 100)) +
  tm_fill(col = "name", alpha = 0.5) +
  tm_layout(
    title = "Source polygons simplified", frame = FALSE,
    legend.show = FALSE)

tmap_arrange(m1, m2, ncol = 2)
```

By contrast if we use the simplification function from `rmapshaper` topology is preserved, even with pretty dramatic simplification (here retaining only 25% of polygon vertices.)

```{r fig.width=8, fig.height=5}
library(rmapshaper)

m3 <- tm_shape(polygons %>% ms_simplify(keep = 0.25)) +
  tm_fill(col = "name", alpha = 0.5) +
  tm_layout(
    title = "Topology preserved", frame = FALSE,
    legend.show = FALSE)

tmap_arrange(m1, m3, ncol = 2)
```

The `rmapshaper::topojson_write` function also allows you to write your polygons out to the topology preserving [TopoJSON format](https://github.com/topojson/topojson).

`rmapshaper` is associated with the excellent [MapShaper tool](http://www.mapshaper.org/).

```{r}
#| code-fold: true
# License (MIT)
#
# Copyright (c) 2023 David O'Sullivan
#
# Permission is hereby granted, free of charge, to any person
# obtaining a copy of this software and associated documentation
# files (the "Software"), to deal in the Software without restriction,
# including without limitation the rights to use, copy, modify, merge,
# publish, distribute, sublicense, and/or sell copies of the Software,
# and to  permit persons to whom the Software is furnished to do so,
# subject to the following conditions:
#
# The above copyright notice and this permission notice shall be included
# in all copies or substantial portions of the Software.
#
# THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS
# OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
# FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL
# THE AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
# LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING
# FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
# DEALINGS IN THE SOFTWARE.
```

