---
title: "Figure 3.10 Raster aggregation and disaggregation"
lightbox: true
categories: 
  - figures
  - code
  - R
execute: 
  cache: true
  freeze: auto
knitr:
  opts_chunk: 
    warning: false
    message: false
---

This figure shows how you can't recover raster information after aggregation.

```{r message=FALSE}
library(terra)
library(tmap)
library(dplyr)
```

Get the data&mdash;you'll need some of your own data here.

```{r}
z <- rast("raster-data.tif")
```

Aggregate and disaggregate by 10

```{r}
z10 <- z %>% 
  aggregate(10)

z_dash <- z10 %>%
  resample(z)
```

Make maps. For the images in the book, I dissolved the raster layers to polygons to create smaller SVG outputs that I was better able to control in making final graphics. For online, I don't need to worry about that.

```{r}
m1 <- tm_shape(z) + 
  tm_raster(breaks = seq(180, 280, 10), 
            palette = "BrBG") + 
  tm_layout(frame = FALSE, legend.show = FALSE)

m2 <- tm_shape(z10) + 
  tm_raster(breaks = seq(180, 280, 10), 
            palette = "BrBG") + 
  tm_layout(frame = FALSE, legend.show = FALSE)

m3 <- tm_shape(z_dash) + 
  tm_raster(breaks = seq(180, 280, 10), 
            palette = "BrBG", style = "fixed") + 
  tm_layout(frame = FALSE, legend.show = FALSE)

tmap_arrange(m1, m2, m3)
```

```{r}
#| code-fold: true
# License (MIT)
#
# Copyright (c) 2023 David O'Sullivan
#
# Permission is hereby granted, free of charge, to any person
# obtaining a copy of this software and associated documentation
# files (the "Software"), to deal in the Software without restriction,
# including without limitation the rights to use, copy, modify, merge,
# publish, distribute, sublicense, and/or sell copies of the Software,
# and to  permit persons to whom the Software is furnished to do so,
# subject to the following conditions:
#
# The above copyright notice and this permission notice shall be included
# in all copies or substantial portions of the Software.
#
# THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS
# OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
# FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL
# THE AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
# LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING
# FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
# DEALINGS IN THE SOFTWARE.
```