---
title: "Figure 3.9: Equal area world in a square"
categories: 
  - figures
  - code
  - R
execute: 
  cache: true
  freeze: auto
knitr:
  opts_chunk: 
    warning: false
    message: false
---

This equal-area projection could replace Web Mercator as the top level global web tile 'world in a square'.

With astonishing foresight, it was proposed for that purpose in this paper in 1986:

> Tobler WR and ZT Chen. 1986. A quadtree for global information storage. _Geographical Analysis_ **18**(4) 360–371. doi: [10.1111/j.1538-4632.1986.tb00108.x](https://dx.doi.org/10.1111/j.1538-4632.1986.tb00108.x).

Given the upset that accompanied the Gall-Peters projection, another 'odd-looking' equal-area projection (only 'odd' because people are so accustomed to Mercator), it's not clear it would catch on! See this paper for a discussion of that controversy:

> Crampton, J. 1994. Cartography’s defining moment: the Peters projection controversy, 1974–1990. _Cartographica_ **31**(4) 16–32. doi: [10.3138/1821-6811-L372-345P](https://dx.doi.org/10.3138/1821-6811-L372-345P).

In any case, this is simply a standard cylindrical equal-area projection with standard parallels chosen to make the whole map area square. 

```{r message=FALSE}
library(sf)
library(tmap)
library(tmaptools)
library(dplyr)
```

Use the supplied `World` dataset (in lat-lon EPSG 4326) and project as required!

```{r fig.width=10, fig.height=9}
data("World") 
world_tc <- World %>%
  select(geometry) %>%
  st_transform("+proj=cea lat_ts=55.654")

tm_shape(world_tc, bbox = bb(xlim = c(-180, 180),
                             ylim = c(-90, 90))) +
  tm_fill(col = "#aaeecc") +
  tm_graticules(x = seq(-180, 180, 15), y = seq(-90, 90, 15),
                labels.show = FALSE, lwd = 0.5) +
  tm_layout(inner.margins = 0, bg.color = "#ddeeff")
```

```{r}
#| code-fold: true
# License (MIT)
#
# Copyright (c) 2023 David O'Sullivan
#
# Permission is hereby granted, free of charge, to any person
# obtaining a copy of this software and associated documentation
# files (the "Software"), to deal in the Software without restriction,
# including without limitation the rights to use, copy, modify, merge,
# publish, distribute, sublicense, and/or sell copies of the Software,
# and to  permit persons to whom the Software is furnished to do so,
# subject to the following conditions:
#
# The above copyright notice and this permission notice shall be included
# in all copies or substantial portions of the Software.
#
# THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS
# OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
# FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL
# THE AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
# LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING
# FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
# DEALINGS IN THE SOFTWARE.
```
